---
title: "M01-2-Introduction to Literate Programming-Application Assignment"
author: "Alexa Bueno"
date: '`r format(Sys.time())`'
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-line-numbers: true
    code-fold: false
    code-link: true
    embed-resources: true
editor: visual
execute:
  freeze: auto
  warning: false
---

# M01-2 Application Assignment

## Ex 1

> Create a variable called X and assign the text "This is my first assignment" to it.

```{r}
X <- "This is my first assignment"
X
```

## Ex 2: Adding Texts in Base R

### Ex 2.1 Learning `paste()`

::: callout-tip
I used a few examples to see how `paste()` and `paste0()` work differently.
:::

```{r}
paste("Hello", "world")
paste("Hello", "world", sep = "-")
paste0("Hello", "world")
```

### Ex 2.2 Adding text to `X`

```{r}
X <- paste(X, "and I'm loving it!")
X
```

## Ex 3

> Create a vector called Y with the numbers 2, 3, 4, and 5. Then multiply the vector by 2.

```{r}
Y <- c(2, 3, 4, 5)
Y <- Y * 2
Y
```

## Ex 4

> Print both the variable `X` and the vector `Y`.

```{r}
X
Y
```

## Ex 5

> Show the maximum and minimum value of `Y`.

```{r}
max(Y)
min(Y)
```

## Ex 6

> Load the `ggplot2` package and look at the first six rows of `economics`.

```{r}
# install.packages("ggplot2")
library(ggplot2)

head(economics)
```

## Ex 7

### Understanding the `economics` Data

The `economics` data has six variables:

- `date`: the month and year of the observation
- `pce`: personal consumption expenditures
- `pop`: total population
- `psavert`: personal savings rate
- `uempmed`: median duration of unemployment
- `unemploy`: number of unemployed people

I decided to compare `unemploy` and `uempmed` because I wanted to see if more unemployment is connected to people staying unemployed longer.

```{r}
plot <- ggplot(economics, aes(x = unemploy, y = uempmed)) +
  geom_point()

plot +
  labs(
    title = "Unemployment and Duration of Unemployment",
    x = "Number of Unemployed People",
    y = "Median Duration of Unemployment"
  )
```

## Ex 8

> I recreated the chart using the pipe operator.

```{r}
#| label: fig-unemployment
#| fig-cap: "Scatterplot showing unemployment and median duration of unemployment."
#| fig-width: 5
#| fig-asp: 0.6

economics |>
  ggplot(aes(x = unemploy, y = uempmed)) +
  geom_point() +
  labs(
    title = "Unemployment and Duration of Unemployment",
    x = "Number of Unemployed People",
    y = "Median Duration of Unemployment"
  )
```

## Ex 9

::: callout-note
In @fig-unemployment, it looks like there is a positive relationship between unemployment and median unemployment duration. When the number of unemployed people is higher, the median amount of time people stay unemployed also seems to be higher.
:::

## Ex 10

::: callout-important
I tried to organize this codebook so it would be easier to read later.
:::

### What I used in this report

1.  I used headings with `#`, `##`, and `###`.
2.  I used **bold text** for emphasis.
3.  I used *italicized text* in the report.
4.  I used `callout-tip` for a hint/note.
5.  I used `callout-note` to explain my finding.
6.  I used `callout-important` for an important organization note.
7.  I used blockquotes for some task instructions.
8.  I used backticks around R functions like `paste()` and `ggplot()`.
9.  I cross-referenced the figure with @fig-unemployment.
10. I used `#|` options inside a code chunk.
11. I added a table of contents in the YAML header.
12. I used numbered sections in the YAML header.
13. I added a figure caption.
14. I used separate code chunks for each question.
15. I used bullet points to explain the variables.

Overall, this assignment helped me practice putting code, notes, and output together in one Quarto document.